---
title: "BCC Statistics:</br>Companion Course"
author: Gerardo Pelayo García
output: rmdformats::readthedown

---

<!--

Package for format of html document

install.packages("rmdformats")

---

Copy and paste for code chunks; not available on CRAN

install.packages("remotes")
remotes::install_github("rlesur/klippy")
-->

```{r set_up, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

Copy and paste the code blocks on this webpage to *R Studio*, editing specific parts the code to fit your data.</br></br>

---

# 0. Introduction

## Getting started

### *Sign up on R Studio Cloud*

Make an account on <a href="https://rstudio.cloud" target="_blank">RStudio Cloud</a>. Use your Argosy Gmail account to make signing-in easier.
</br></br>

### *Create an R file to save work*

1. Sign into <a href="https://rstudio.cloud" target="_blank">RStudio Cloud</a>
</br></br>
2. Click *New Project* > *New RStudio Project*
</br></br>
3. Click *File* > *New File* > *R Script*
</br></br>
4. Click the save icon and name your file.
</br></br>
5. *(optional)* Click *Tools* > *Global Options*. You can edit *Appearance* (colors and themes) and *Pane Layout* (where the 4 windows are placed) to customize the user interface. Click *Apply* to save your changes.</br></br>

### *Install the packages in this guide*

Copy the code block below and paste it into your *R file*, then click *Source*. The initial download may take a few minutes.</br>

After a package is installed, the same code below will load the packages. It should always be at the top of the *R file* before any other code appears.

```{r, eval=FALSE}
packages = c('DT', 'dplyr', 'magrittr', 'kableExtra', 'plotly')

package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```
</br></br>

## How to use this guide

* Each code block has a copy-to-clipboard button on the top left. 
  + Click this button and paste it into *R Studio Cloud*. 
  + To save and organize your work, always use an *R file* and click *Source* to runn the code.
  + Always paste new code underneath previous code, never above it (unless you understand what you are doing).
  + If you paste the code into the *Console*, press enter.
</br></br>
* Use the *table of contents* and its sub-headers to jump to different sections.
</br></br>
* Every list (of numbers) and *data frame* (table with data) is named ```df```.
  + Every function and example references this variable consistently. 
  + Each section will explain what ```df``` should be in each case, including what type of data each column should be.
</br></br>

## Data used in examples

For a `list` of numbers, the data `df` used in each example is always:

```{r, eval=FALSE}
df <- c(1, 2, 3, 4, 5)
```

For a `table` with 2 or more columns, the data used in each example is:

```{r, eval=FALSE}
df <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c(6, 7, 8, 9, 10)
)
```

For advanced figures, the data used for the examples is pre-installed in *R Studio*. The data refers to different car models, which is shown below:

```{r}
df <- mtcars
df$car <- rownames(df)
df <- df[,c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)]

DT::datatable(df, 
  caption = 'Pre-installed data set \'mtcars\'',
  rownames = FALSE)
```
</br></br>

<!--
## Installing packages

### *Default method*

To install *packages*, additional code that allows *R* to do more advanced tasks, we need to use two different functions.

```{r, eval=FALSE}
# DO NOT RUN
install.packages('name_of_package')
```

Paste the code an *R file* and run it once by clicking *Source*, then place a pound sign (```#```) before it in the same row to prevent *R* from trying to install it a second time. This makes it easier for someone else to see packages they need to run your code.

Names of packages are case-sensitive and never include spaces or special characters.

If you ever see a pop-up saying a package is required and asking if you want to install it, **always click 'Yes' to save time**. Any downloads through *R Studio Cloud* are always safe.
</br></br>

### *Via GitHub*

In some cases, packages are only available through <a href="https://github.com" target="_blank">GitHub</a>. If this is the case, first check if the *devtools* package is installed by pasting this into the console:

```{r, eval=FALSE}
library(devtools)
```

If you receive an error message, paste this into an *R file* and click *Source* before typing ```#``` before it in the same row:

```{r, eval=FALSE}
install.packages('devtools')
```

Afterwards, find the name of the developer and the name of their project. Without spaces, surrounded by quotes, and separated by a slash (```'name_of_developer/name_of_project'```), paste this into the console:

```{r, eval=FALSE}
devtools::install_github('name_of_developer/name_of_project')
```

Double-check for spelling, capital and lower-case letters, and surrounding quotes.

Be sure to include whatever functions were necessary in your *R file* with a preceding ```#```.
</br></br>
-->

---

# 1. Write data

After pasting the code into an *R file*, replace the comma-separated numbers with your data. Only the numbers are the commas between them should be replaced.
</br></br>

### List of numbers

The data are provided as a list of numbers without any variables, or a table with only one column. In other words, only one thing is being measured.

```{r}
df <- c(1, 2, 3, 4, 5)
```
</br></br>

### *Table with 2 columns*

```{r}
df <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(6, 7, 8, 9, 10)
)
```
</br></br>

### *Table with 3 columns*

```{r}
df <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(6, 7, 8, 9, 10),
  'z' = c(11, 12, 13, 14, 15)
)
```

### *Table with 4 or more columns*

Follow the pattern in the previous examples

### *Name the columns something else*

Replace the letters *x*, *y*, and *z* (if relevant) with the names, surrounded by quotation marks. In general, do not include spaces or special characters in the name, excluding dashes (-), underscores (_), and numbers(0-9). Note that these names are case-sensitive (**n**ame and **N**ame are different).

```{r, eval=FALSE}
# DO NOT RUN

df <- data.frame(
  'good_name' = c(),
  'good-name1' = c(),
  'goodName' = c(),
  
  'bad name' = c(),
  'Bad.Name' = c(),
  'Really Bad Name' = c()
)
```
</br></br>

---

# 2. Calculate statistics

```{r}
# List
df_list <- c(1, 2, 3, 4, 5)

# Table with 2 columns
df_table <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(6, 7, 8, 9, 10)
)
```

## Measures of central tendency

**IMPORTANT - PLEASE READ.**

For the code below, ```x``` is the name of the column from the table ```df``` you are referencing.

```{r, eval=FALSE}
df <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(6, 7, 8, 9, 10)
)
```

If the data is a list, the name of the list is ```df```.

```{r, eval=FALSE}
df <- c(1, 2, 3, 4, 5)
```
</br></br>

```{r, echo=FALSE}
df <- c(1, 2, 3, 4, 5)
```

### *Mean/Average*

The output (3) will appear in the console.

**List ```df```**

```{r, eval=FALSE}
mean(df)
```

**Table ```df``` with column ```x```**

```{r, eval=FALSE}
mean(df$x)
```

```{r, echo=FALSE, results='asis'}
cat('= ', mean(df))
```
</br></br>

### *Median*

The output (3) will appear in the console.


**List ```df```**

```{r, eval=FALSE}
median(df)
```

**Table ```df``` with column ```x```**

```{r, eval=FALSE}
median(df$x)
```

```{r, echo=FALSE, results='asis'}
cat('= ', median(df))
```
</br></br>

### *Mode*

**_See custom functions_.**
</br></br>

## Measures of dispersion

### *Variance*

**List ```df```**

```{r, eval=FALSE}
var(df)
```

**Table ```df``` with column ```x```**

```{r, eval=FALSE}
var(df$x)
```

```{r, echo=FALSE, results='asis'}
cat('= ', var(df))
```
</br></br>

### *Standard deviation*

**List ```df```**

```{r, eval=FALSE}
sd(df)
```

**Table ```df``` with column ```x```**

```{r, eval=FALSE}
sd(df$x)
```

```{r, echo=FALSE, results='asis'}
cat('= ', sd(df))
```
</br></br>

### *Standard error*

**_See custom functions_.**
</br></br>

## Custom functions

Copy and paste the code below into an *R file*, then click *Source*. To use the function, enter the code below the pasted code or into the console.

### *Mode*

```{r}
# Sort by frequency, descending order (child: mode())
sort_mode <- function(x) {
  temp <- data.frame(table(x))
  temp <- temp[order(-temp$Freq),]
  
  rownames(temp) <- NULL
  names(temp) <- c('value', 'frequency')
  
  return(temp)
}

# Find the mode (parent: sort_mode())
mode <- function(x) {
  x <- sort_mode(x)
  rows_x <- nrow(x)
  max_freq <- max(x$frequency)
  
  x$is_max <- 0
  x$is_max[x$frequency==max_freq] <- 1
  
  x <- x[x$is_max==1,]
  x <- data.frame(as.numeric(as.character(x$value)))
  rownames(x) <- NULL
  
  if(nrow(x)==rows_x){
    x = 'no mode'
  } else {
    names(x) <- c('mode(s):')
  }
  
  return(x)
}
```

```{r}
mode(df)
```

### *Manually calculate r, r^2, a (y-intercept), and b (slope)*

```{r}
df <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(6, 7, 8, 9, 10)
)
```

```{r}
# Create table to manually calculate several statistics (r, r^2, a, and b)
make_table <- function(df) {
  require(DT)
  
  df$xy <- df$x*df$y
  df$x2 <- df$x**2
  df$y2 <- df$y**2
  
  all_sums <- c(sum(df$x), sum(df$y), sum(df$xy), sum(df$x2), sum(df$y2))
  df <- rbind(df, all_sums)
  rownames(df)[rownames(df)==as.character(nrow(df))] <- 'Total Sums'
  table <- DT::datatable(df,
    extensions = 'Buttons',
    caption = paste('Total observations: n = ', nrow(df)-1),
    options = list(
      dom = 'Bt',
      buttons = c('copy', 'csv', 'excel')
  ))
  return(table)
}
```

```{r}
make_table(df)
```

### *Standard error*

```{r}
df <- c(1, 2, 3, 4, 5)
```

```{r}
se <- function(x) {
  temp <- round(sd(x)/sqrt(length(x)), digits=4)
  
  return(temp)
}
```


```{r}
se(df)
```

### *Manually calculate continuius probabilites*

**1. Write the data.**

```{r}
df <- data.frame(
  'x' = c(1, 2, 3, 4, 5),
  'y' = c(0.1, 0.2, 0.3, 0.4, 0.5)
)
```


**2. Copy and paste the function below. If an *R file* is being used, click *Source*.**

```{r}

# Create table to manually calculate continuous probabilities
probability_table <- function(df) {
  df$mean <- df$x*df$y
  mean <- sum(df$mean)
  df <- subset(df, select=-c(mean))
  df$`x-m` <- df$x - mean
  df$`(x-m)^2` <- round(df$`x-m`**2, digits=3)
  df$`(x-m)^2*p(x)` <- round(df$`(x-m)^2`*df$y, digits=3)
  names(df)[names(df)=='y'] <- 'p(x)'
  
  var <- round(sum(df$`(x-m)^2*p(x)`), digits=3)
  sd <- round(sqrt(sum(df$`(x-m)^2*p(x)`)), digits=3)
  var_row <- c('', '', '', '', var)
  df <- rbind(df, var_row)
  sd_row <- c('', '', '', '', sd)
  df <- rbind(df, sd_row)
  rownames(df)[rownames(df)==as.character(nrow(df)-1)] <- 'Var.'
  rownames(df)[rownames(df)==as.character(nrow(df))] <- 'Std. Dev.'
  
  table <- DT::datatable(df,
    extensions = 'Buttons',
    caption = paste('See bottom rows for Variance (Var.) 
                    and Standard Deviation (Std. Dev.)'),
    options = list(
      dom = 'Bt',
      buttons = c('copy', 'csv', 'excel')
  ))
  return(table)
}
```

**3. Call the function `probability_table()` with the data `df` inside the parentheses.**

```{r}
probability_table(df)
```
</br></br>

---

# 3a. Make basic tables

```{r, echo=FALSE}
# Change data frame df to pre-installed code for examples
df <- mtcars
```

**For data with 2 or more columns only. Lists will not work.**
</br></br>

## Standard table

Replace *Title* with a title for the table.

```{r}
library(magrittr)

df%>%
  kableExtra::kbl(caption = 'Title')%>%
  kableExtra::kable_styling()
  
```
</br></br>

## Other themes

Replace *Title* with a title for the table.

```{r}
kableExtra::kbl(cbind(df, df)) %>%
  kableExtra::kable_paper() %>%
  kableExtra::scroll_box(width = "100%", height = "200px")
```

```{r}
library(magrittr)

df %>%
  kableExtra::kbl(caption = 'Title') %>%
  kableExtra::kable_paper("hover")
```

```{r}
library(magrittr)

df %>%
  kableExtra::kbl(caption = 'Title') %>%
  kableExtra::kable_paper("hover", full_width = F)
```

**Additional information can be found <a href="https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html" target="_blank">here</a>.**
</br></br>

---

# 3b. Make interactive tables

**For technical purposes only.**

The pre-installed data set *mtcars* has the car names as the row names instead of rows of data. By default, row names in a table are just the row number (first row = 1, second row = 2, etc.). The code below adds the row names as a column and reorders the columns, allowing us to keep the car names (first) while removing the row names.

```{r}
df <- mtcars
df$car <- rownames(df)
df <- df[,c(12, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11)]
```
</br></br>

**For data with 2 or more columns only. Lists will not work.**
</br></br>

## Standard table

Replace *Title* with a title for the table.

```{r}
DT::datatable(df, 
  rownames = FALSE,
  caption = paste('Title')
  )
```
</br></br>

## With download buttons

```{r}
DT::datatable(df,
  rownames = FALSE,
  extensions = 'Buttons',
  caption = paste('Title'),
  options = list(
  dom = 'Bt',
  buttons = c('copy', 'csv', 'excel')
  ))
```
</br></br>

## With column filters
```{r}
DT::datatable(df,
  rownames = FALSE,
  extensions = 'Buttons',
  caption = paste('Title'),
  options = list(
  dom = 'Bt'),
  filter = 'top')
```
</br></br>

## Other themes

Replace *Title* with a title for the table.

```{r}
DT::datatable(df, 
  rownames = FALSE,
  caption = paste('Title'),
  class = 'cell-border stripe'
  )
```

```{r}
DT::datatable(df,
  rownames = FALSE,
  extensions = 'Buttons',
  caption = paste('Title'),
  class = 'cell-border stripe',
  options = list(
  dom = 'Bt',
  buttons = c('copy', 'csv', 'excel')
  ))
```
**Additional information can be found <a href="https://rstudio.github.io/DT/" target="_blank">here</a>.**
</br></br>

---

# 4. Statistical figures

The following section references columns within data sets. To reference a specific column for a data set ```df```, we use a dollar sign ```$``` afterwards and write the name of the column:

```{r, eval=FALSE}
# DO NOT RUN
df$name_of_column
```

The column names of the data set used for this example can be found below.

```{r, echo=FALSE}
cat(colnames(df))
```
</br></br>

## Scatter plot

Replace *mpg* and *wt* with the column names of **quantitative** data. Note that each column name is preceded with a tilde ```~``` without a space. Replace *Title* with a title for the figure.

```{r, message=FALSE}
library(magrittr)

plotly::plot_ly(
  data = df,
  x = ~wt, 
  y = ~mpg)%>%
  plotly::layout(title='Title')
```
</br></br>

## Scatter plot with a legend

Replace *mpg* and *wt* with the column names of **quantitative** data. Note that each column name is preceded with a tilde ```~``` without a space. Replace *Title* with a title for the figure.
</br></br>

### *Continuous variable*

Replace *cyl* with a column that will determine the color each point receives. If the data is numerical, it will create a gradient by default.

```{r, message=FALSE}
library(magrittr)

plotly::plot_ly(
  data = df,
  x = ~wt, 
  y = ~mpg,
  color = ~cyl)%>%
  plotly::layout(title='Title')
```

### *Discrete variable*

Although the column *cyl* is numerical, there are only three unique values that appear. By appending ```factor```, the column is treated as qualitative data, where a unique number is a group.
</br></br>

Replace *cyl* with **qualitative data**, including numerical data with discrete values only.

```{r, message=FALSE}
library(magrittr)

plotly::plot_ly(
  data = df,
  x = ~wt, 
  y = ~mpg,
  color = ~factor(cyl))%>%
  plotly::layout(title='Title')
```
</br></br>

## Bar chart

```{r, message=FALSE}
library(magrittr)

plotly::plot_ly(
  data = df,
  x = ~car, 
  y = ~mpg,
  type = 'bar')%>%
  plotly::layout(title='Title')
```
</br></br>

## Pareto chart

```{r, message=FALSE}
library(magrittr)

plotly::plot_ly(
  data = df,
  x = ~reorder(car, -mpg), 
  y = ~mpg,
  type = 'bar')%>%
  plotly::layout(title='Title')%>%
  plotly::layout(xaxis = list(title='car'))
```
</br></br>

## Box-and-whisker plot

```{r}
plotly::plot_ly(data = df, y=~mpg, type = 'box', hoverinfo = 'y', name = '')%>%
  plotly::layout(title = 'Title')
```
</br></br>

## Additional information

**Additional information can be found <a href="https://plotly.com/r/" target="_blank">here</a>.**

---


